---
title: "Programming in R - Day 3, afternoon"
author: "Stacey Borrego"
date: "1/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Packages needed for today

```{r eval=FALSE}
# installations can be `c`ombined
install.packages(c("devtools", "roxygen2", "rmarkdown", "knitr")) 
```

### Best Practices for Writing R Code

* [Best Practices for Writing R Code](http://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/index.html)
* [Style Guide (Advanced R by Hadley Wickham)](http://adv-r.had.co.nz/Style.html)
* Code clean-up helpers:
  * RStudio --> Code --> Reindent Lines
  * RStudio --> Code --> Reformat Code

**Annotation**

* Include description of the code
  * What it does
  * Why it was written
  * The system it was written for
  
```{r}
# This is code to replicate the analyses and figures from my 2014 Science
# paper. Code developed by Sarah Supp, Tracy Teal, and Jon Borelli
```

**Version Control**

* Use version control with clear and helpful commit messages
  * [git](https://swcarpentry.github.io/git-novice/)
  * [RStudio with Git Integration](https://web.archive.org/web/20190103191213/https://support.rstudio.com/hc/en-us/articles/200532077?mobile_site=true)
  
**Dependencies**

* Load all the packages necessary at the top of your code using `library()`
  * [The difference between `library()` and `require()`](https://www.r-bloggers.com/2017/10/loading-r-packages-library-or-require/)
* May want to consider using the format `package::function()` especially if sharing scripts

```{r eval=FALSE}
library(ggplot2)
library(reshape)
library(rcolorbrewer)
```

* Reduce hard-coding input and output files and working directories paths
  * Use variables to assign input and output files
  * Use relative paths from the project directory
* Define variables early on in the script

*Relative path* - Best practice
```{r eval=FALSE}
dat <- read.csv(file = "files/dataset-2013-01.csv", header = TRUE)
```

*Absolute path* - not recommended

```{r eval=FALSE}
dat <- read.csv(file = "/Users/Karthik/Documents/sannic-project/files/dataset-2013-01.csv", header = TRUE)
```

*Good Example*: variables are defined at the top of script and values are calculated using variables
```{r eval=FALSE}

input_file <- "data/data.csv" 
output_file <- "data/results.csv"

# --------------------------------

# read input
input_data <- read.csv(input_file)
# get number of samples in data
sample_number <- nrow(input_data)
# generate results
results <- some_other_function(input_file, sample_number)
# write results
write.table(results, output_file)
````

*Bad Example*: variables are not used when possible or defined at the beginning of script 
```{r eval=FALSE}
# check
input_data <- read.csv("data/data.csv")
# get number of samples in data
sample_number <- nrow(input_data)
# generate results
results <- some_other_function("data/data.csv", sample_number)
# write results
write.table("data/results.csv", output_file)
```

**Identify and segregate distinct components in your code**

* Annotate and mark your code with comments. This will be helpful to understand what the code does, its requirements, and divide sections that perform similar tasks.
  * `#`
  * `#-----------`
* Keep your code in bite-sized chunks. If a single function or loop gets too long, consider looking for ways to break it into smaller pieces.
* Include functions at the top of your script
  * If there are many custom functions, put them in their own script and source the file.
    * Keep source files in the same project directory, then use relative paths.
  
```{r eval=FALSE}
source("scripts/my_genius_fxns.R")
```

**Extra Tips**

* Don’t repeat yourself–automate! If you are repeating the same code over and over (3x), use a loop or a function. 
* Use a consistent style within your code. 
  * Consistency makes code easier to read and problems easier to spot.
  * Example: name all matrices something ending in _mat. 
  * [Style Guide (Advanced R by Hadley Wickham)](http://adv-r.had.co.nz/Style.html)
* Keep track of `sessionInfo()` in your project folder. 
  * Notes all of the packages and versions used in the current project. If a newer version of a package changes the way a function behaves, you can reinstall the version that worked
  * [CRAN](https://cran.r-project.org/), all versions of packages are archived
* R can run into memory issues. R will not clean up unused memory for a while after you delete objects. You can force R to tidy up its memory by using `gc()` (Garbage Collection). See `?gc` for more information
* `rm(list=ls())` can be a useful command, but do not put in your scripts.
  * `rm()` is used to remove all the objects from the workspace when you use `list=ls()` as base
  * instead of using this command, start a new session
    * RStudio --> Session --> New Session

##### Key Points

* Start each program with a description of what it does.
* Load all required packages at the top
* Consider what working directory you are in when sourcing a script
* Use comments to mark off sections of code
* Put function definitions at the top of your file, or in a separate file if there are many.
* Name and style code consistently.
* Break code into small, discrete pieces.
* Keep all of the source files for a project in one directory and use relative paths to access them.
* Keep track of the memory used by your program.
* Always start with a clean environment instead of saving the workspace.
* Keep track of session information in your project folder.
* Have someone else review your code.
* Use version control.

### Dynamic Reports with knitr

* [Dynamic Reports with knitr](http://swcarpentry.github.io/r-novice-inflammation/07-knitr-R/index.html)
* [Book: Dynamic Documents with R and knitr](https://duhi23.github.io/Analisis-de-datos/Yihue.pdf)

* [R markdown cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/)
* [Rmarkdown from RStudio](https://rmarkdown.rstudio.com/)
* [Rmarkdown Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
* [Rmarkdown Basics](https://rmarkdown.rstudio.com/authoring_basics.html)
* [Get Started](https://rmarkdown.rstudio.com/lesson-1.html)

**knitr** is an R package that allows you to organize your notes, code, and results in a single document.

* Keeps your writing and results together and makes it easy to update
* knitr documents are written in simple plain text-like format called markdown (Rmd)
* Outputs: 
  * HTML file
  * PDF (requires LaTeX and [Pandoc](https://pandoc.org/installing.html))
  * Word document ()

```{r eval=FALSE}
install.packages("rmarkdown")
install.packages("knitr")

# to use python
install.packages("reticulate")
````

**Set Up:**

* Install `rmarkdown` and `knitr` packages from CRAN
* You may need to restart RStudio
* RStudio: FILE --> NEW FILE --> R MARKDOWN...

Rmarkdown files contain three types of content:

* **Metadata**: An (optional) YAML header surrounded by ---s
* **Code**: code chunks surrounded by ```s and language indicated by {}
  * Command+Option+I or Ctrl+Alt+I
* **Text**: mixed with simple text formatting

```{bash}
pwd
ls /Users/stacey/Desktop/shell-lesson-data/*.txt
```

```{r}
# install.packages("reticulate")
library(reticulate)
knitr::knit_engines$set(python = reticulate::eng_python)
use_python("/Library/Frameworks/Python.framework/Versions/3.10/bin/python3")
```

```{python eval=FALSE}
```

### Making Packages in R

* [Making Packages in R](http://swcarpentry.github.io/r-novice-inflammation/08-making-packages-R/index.html)
* [R Packages by Hadley Wickham and Jenny Bryan](https://r-pkgs.org/)
* [rOpenSci](https://devguide.ropensci.org/index.html)


An R package is the basic unit of reusable code. If you want to reuse code later or want others to be able to use your code, you should put it in a package.

An R package requires four components:

* **DESCRIPTION** file with metadata about the package 
* **R** directory with the code
* **man** "man"ual or documentation
* **NAMESPACE** file listing user-level functions in the package
* Optional components

Easily prepare packages using devtools!

```{r eval=FALSE}
install.packages(c("devtools", "roxygen2"))

# Load packages to make help make your package
library("devtools")
library("roxygen2")

# Move to location that package will be saved to 
getwd()
setwd("/Users/stacey/Desktop")

# This will create a new project in a new session
create_package("tempConvert")

# Reload packages
library("devtools")
library("roxygen2")

# Save each function in a script with descriptions written in the roxygen format

# SCRIPT --------------------------------------------
<!-- #' Converts Fahrenheit to CelSsius -->
<!-- #' -->
<!-- #' This function converts input temperatures in Fahrenheit to Celsius. -->
<!-- #' @param temp_F The temperature in Fahrenheit. -->
<!-- #' @return The temperature in Celsius. -->
<!-- #' @export -->
<!-- #' @examples -->
<!-- #' fahrenheit_to_kelvin(32) -->
# There are even more options

# Write function HERE
  
# SCRIPT --------------------------------------------
  
# Make documentation
document()

# Test
# In project session Build --> Clean and Rebuild 
# Control+Shift+B
?tempConvert
celsius_to_kelvin(2)


# Open a new session and load your package
# Session --> New Session
library(tempConvert)
```

* [Put your R package on Github](https://kbroman.org/pkg_primer/pages/github.html)
* Just for fun and an example, here is a package I have on Github ---> [saguaRo](https://github.com/sborrego/saguaRo)


### Command Line Programs

* [Command Line Programs](http://swcarpentry.github.io/r-novice-inflammation/05-cmdline/index.html)
* [R Basic Operations](http://swcarpentry.github.io/r-novice-inflammation/reference.html#standard-input-stdin)

Making an R script that can be run on the command line in the following scenarios:

* If no filename is given on the command line, read data from standard input.
* If one or more filenames are given, read data from them and report statistics for each file separately.
* Use the --min, --mean, or --max flag to determine what statistic to print.

To make this work, we need to know how to handle command-line arguments in a program, and how to get at standard input. We’ll tackle these questions in turn below.


